---
layout: post
title: "R Cool Functions List"
date: "July 24, 2015"
categories: R
---

* TOC
{:toc}

```{r, echo = FALSE}
library(jn.general)
lib(data, viz)
```

# Links

* [R reference][r_ref]{:target = "_blank"}
* [dplyr cheatsheet][dplyr_ref]{:target = "_blank"}
* [data.table cheatsheet][data.table_ref]{:target = "_blank"}

[r_ref]: https://drive.google.com/file/d/0B5VF_idvHAmMMWN5dmhaT05IRkk/view?usp=sharing
[dplyr_ref]: https://drive.google.com/file/d/0B5VF_idvHAmMblBxTjEwRWZXYjQ/view?usp=sharing
[data.table_ref]: https://drive.google.com/file/d/0B5VF_idvHAmMOTVOSnRVMkFHczQ/view?usp=sharing

The `broom` package has three main functions that deal with a variety of modeling functions

* `tidy`: extracts the model estimates and p-values
* `glance`: extracts the model statistics ($$R^2$$, etc)
* `augment`: extracts the individual data point statistics (fitted values, residuals, leverages, etc)

# Numeric Types
```{r, echo = FALSE}
func <- c("seq(from, to, by)",
          "range()",
          "quantile(x, probs)",
          "cut(x, breaks)",
          "dist(x)"
          )

descr <- c("generate regular sequences",
           "produces sample quantiles",
           "computes min/max of all args",
           "cuts up continuous values into factor intervals",
           "computes pairwise distances for all points"
           )

data.frame(func, descr) %>% kable
```

# Logical Types
```{r, echo = FALSE}
func <- c("which()",
          "%in%", 
          "between(x, lower, upper)"
          )

descr <- c("gives the TRUE indices",
           "finds matches", 
           "whether x is between l and u"
           )

data.frame(func, descr) %>% kable
```

# Character Types
```{r}
func <- c(""
          )

descr <- c(""
           )

data.frame(func, descr) %>% kable
```

# Input/Output
```{r}
func <- c("list.files(dir)",
          "save(list = ls(), file = '__.Rdata')"
          )

descr <- c("obtain a list of all files in given file path",
           "save all vars for another session"
           )

data.frame(func, descr) %>% kable
```


# Any Type
```{r}
func <- c(""
          )

descr <- c(""
           )

data.frame(func, descr) %>% kable
```
sample (takes a random sample of x), order, sort, table, str, summary, names, colnames, rownames, assign("name", obj), get("name"), expand.grid, interaction, mapvalues