---
layout: post
title: "Subsetting With dplyr (R)"
date: "August 1, 2015"
categories: ['r programming']
---

* TOC
{:toc}

```{r global_opts, echo = FALSE, message = FALSE, warning = FALSE}
library(jn.general)
lib(data, viz)
```

The dplyr package provides handy tools to slice and dice data into various formats with ease.   

# Example Data
Let's use the random generator function, `jn.general::rdata()`, to generate an example data set.

```{r generate random data}
set.seed(1)
data <- jn.general::rdata(n = 750,
              # posttests
              gen_seq(0, 100, length.out = 40, name = "posttest_score"),
              gen_subject(name = "posttest_subject"),
              gen_seq(10, 12, name = "posttest_grade"),
              # pretests
              gen_char(name = "pretest_7_name", values = "ACT"),
              gen_char(name = "pretest_8_name", values = "Math222"),
              gen_char(name = "pretest_9_name", values = "SAT"),
              gen_char(name = "pretest_10_name", values = "CompSci367"),
              gen_char(name = "pretest_11_name", values = c("ACT", "SAT")),
              gen_seq(0, 100, length.out = 40, ncol = 5, name = paste0("pretest_", 7:11, "_score"), add.na = TRUE),
              gen_seq(8, 11, ncol = 5, name = paste0("pretest_", 7:11, "_grade")),
              gen_char(ncol = 5, name = paste0("pretest_", 7:11, "_version"), values = LETTERS[1:3]),
              # demographics
              gen_gender(name = "d_gender"),
              gen_bool(ncol = 4, name = c("d_black", "d_hispanic", "d_asian", "d_native"), add.na = TRUE, probs = c(0.15, 0.8, 0.05)),
              gen_bool(ncol = 1, name = "d_gifted", probs = c(0.25, 0.25, 0.5), add.na = TRUE), 
              # student id
              .id = FALSE
              )
data %<>% group_by(posttest_subject) %>% mutate(student_id = 1:n()) %>% data.frame
# formatting
data %<>% select(student_id, matches("posttest"), matches("pretest"), matches("d_"))
```

<div class = "dftab">
```{r, echo = FALSE}
data %>% nhuyhoa_df_print()
```
</div><p></p>

# dplyr::slice() and dplyr::filter()
There are a number of functions that allow for rowise manipulation in dplyr, two of which are `dplyr::filter()` and `dplyr::slice()`. 

## Subsetting rows by conditional expression
The `dplyr::filter()` function takes logical conditions as arguments. Combine multiple conditions with `&` (and) or `|` (or). This function is equivalent to base R's subset function.

{% highlight r %}
# subset to students with posttest math scores over 75 and are in the 10th grade
data %>% filter(posttest_grade == 10, posttest_subject == "math", posttest_score > 75) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr filter ex, echo = FALSE}
data %>% filter(posttest_grade == 10, posttest_subject == "math", posttest_score > 75) %>% nhuyhoa_df_print()
```
</div><p></p>

## Subsetting rows by row position
The `dplyr::slice()` function allows rows to be extracted by position.

{% highlight r %}
# extract the 25th-27th students in the data set
data %>% slice(25:27)
{% endhighlight %}
<div class = "dftab">
```{r dplyr slice ex, echo = FALSE}
# extract the 25th-27th students in the data set
data %>% slice(25:27) %>% nhuyhoa_df_print(head = 6)
```
</div><p></p>

# dplyr::select()
The `dplyr::select()` function selects columns from a data set. To deselect columns, place a `-` before the name or special function call. 

## Select columns by name
Selecting columns by name by writing out the name of the column(s), separated by commas.

{% highlight r %}
# select multiple columns by name
data %>% dplyr::select(posttest_score, posttest_grade, d_gender, posttest_subject) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select ex1, echo = FALSE}
data %>% dplyr::select(posttest_score, posttest_grade, d_gender, posttest_subject) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

{% highlight r %}
# deselect multiple columns by name
data %>% dplyr::select(-posttest_score, -posttest_grade, -posttest_subject) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select ex2, echo = FALSE}
data %>% dplyr::select(-posttest_score, -posttest_grade, -posttest_subject) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

## Select columns by numeric position or column name

Columns can also be extracted from a data.frame using a vector of column names or positions. 

For this example, select the demographics columns corresponding to race. 

For numeric position values, just pass the vector as an argument to the select function.

{% highlight r %}
# select based on column position
var_pos <- 25:29 
data %>% dplyr::select(var_pos) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select ex3, echo = FALSE}
var_pos <- 26:29 
data %>% dplyr::select(var_pos) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

For character names, wrap the helper function `one_of()` around the character vector and pass this to the select function.

{% highlight r %}
# select based on column name
var_names <- c("d_black", "d_hispanic", "d_asian", "d_native")
data %>% dplyr::select(one_of(var_names)) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select ex4, echo = FALSE}
var_names <- c("d_black", "d_hispanic", "d_asian", "d_native")
data %>% dplyr::select(one_of(var_names)) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

## Select columns using regex

Helper functions such as `matches()`, `starts_with()`, and `ends_with()` are perhaps the most powerful feature of the select function. 

Start with extracting the pretest score columns. 

Notice that this method is generic (any pretest number it will qualify) and reuseable. 

{% highlight r %}
# use matches to find all pretest score columns
data %>% dplyr::select(matches("pretest_\\d+_score")) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select ex7, echo = FALSE}
data %>% dplyr::select(matches("pretest_\\d+_score")) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

The helper functions `starts_with()` and `ends_with()` are similar to `matches()` but is specialized in that it only looks at the beginning or the end of the string for a pattern match.

Use `starts_with()` and `ends_with()` to extract (1) all posttest information and (2) all test score columns.

{% highlight r %}
# extract all posttest information
data %>% dplyr::select(starts_with("posttest")) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select ex8, echo = FALSE}
data %>% dplyr::select(starts_with("posttest")) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

{% highlight r %}
# extract all test score columns
data %>% dplyr::select(ends_with("_score")) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select ex9, echo = FALSE}
data %>% dplyr::select(ends_with("_score")) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

For added functionality, the `matches()`, `starts_with()`, and `ends_with()` function works even when the regex string is stored as a variable. This tends to be useful when developing functions that runs on any data set, where the regex string can be passed as a parameter. 

## Combining selection techniques

{% highlight r %}
# select multiple columns
data %>% dplyr::select(student_id, matches("pretest_\\d+_score"), d_gender, starts_with("d_")) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select ex10, echo = FALSE}
data %>% dplyr::select(student_id, matches("pretest_\\d+_score"), d_gender, starts_with("d_")) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

{% highlight r %}
# select all pretest info and then deselect pretest 10 info
no_good <- paste0("pretest_10_", c("name", "score", "version", "grade"))
data %>% dplyr::select(student_id, starts_with("pretest_"), -one_of(no_good)) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select ex11, echo = FALSE}
no_good <- paste0("pretest_10_", c("name", "score", "version", "grade"))
data %>% dplyr::select(student_id, starts_with("pretest_"), -one_of(no_good)) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

With very wide data, `dplyr::select()` can be used to reorganize columns. 

{% highlight r %}
# shuffle up column positions
order <- sample(1:ncol(data))
unorganized_data <- dplyr::select(data, order)

# organize columns: id, then posttest info, pretest info, demographics info
unorganized_data %>% dplyr::select(student_id, matches("posttest"), matches("pretest"), d_gender, d_gifted, everything()) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select reorder1, echo = FALSE}
set.seed(1)
order <- sample(1:ncol(data))
unorganized_data <- dplyr::select(data, order)
unorganized_data %>% dplyr::select(student_id, matches("posttest"), matches("pretest"), d_gender, d_gifted, everything()) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

When reorganizing columns if the selectors are too specific, data may be lost in the process. The helper function `everything()` selects all variables not already selected. Including `everything()` at the end of the selection specifications ensures that columns that aren't accounted for in the ordering are included in the returned data frame. 

To move columns to the end of the list, deselect the column and then include `everything()`. 

{% highlight r %}
# move demographic information to the furthest right
unorganized_data %>% dplyr::select(-starts_with("d_"), everything()) %>% head
{% endhighlight %}
<div class = "dftab">
```{r dplyr select reorder2, echo = FALSE}
unorganized_data %>% dplyr::select(-starts_with("d_"), everything()) %>% nhuyhoa_df_print(head = 3)
```
</div><p></p>

# Challenge round
The power of dplyr can be seen with a more challenging problem. 

Clean up this data set by 

* removing  all pretest information that does not correspond to the ACT or SAT
* keep information regarding to students who took a math/reading posttest in the 12th grade

```{r challenge, warning = FALSE}
# grade and subject query of interest
grade_query <- 12
subject_query <- c("math", "read")

# obtains pretest number corresponding to pretests of the ACT and SAT category
pretest_numbers <- data %>% 
  # select all pretest columns
  dplyr::select(matches("pretest_\\d+_name")) %>% 
  # obtain unique test names per column
  apply(2, unique) %>% 
  # filter to elements that contain either ACT or SAT
  jn.general::refine(function(x) any(str_detect(x, "ACT|SAT"))) %>% 
  # obtain the names
  names %>% 
  # extract the pretest number
  str_extract("\\d+") 

# generates the names of the pretest information 
## matches doesn't allow multiple matches so we have to regenerate the names of the pretests information
pretests_of_interest <- paste0("pretest_", pretest_numbers) %>% 
  # interact the pretest names with the values name score and grade
  interaction(c("name", "score", "grade", "version"), sep = "_") %>% 
  # extract the unique names
  levels

# selects posttest, ACT/SAT pretests, demographics and subsets by the grade/subject of interest
final <- data %>% 
  dplyr::select(student_id, starts_with("posttest_"), one_of(pretests_of_interest), starts_with("d_")) %>% 
  dplyr::filter(posttest_grade %in% grade_query, posttest_subject %in% subject_query)
```

<div class = "dftab">
```{r, echo = FALSE}
final %>% nhuyhoa_df_print()
```
</div><p></p>

Note: The `jn.general::refine()` function used here is a wrapper function for `Filter()`. 


[regex_link]: http://jnguyen92.github.io/nhuyhoa//2015/07/Regular-Expressions.html

