---
layout: post
title: "Regular Expressions"
date: "July 25, 2015"
categories: ['data wrangling']
---

* TOC
{:toc}

```{r global_opts, echo = FALSE, message = FALSE, warning = FALSE}
library(jn.general)
lib(data, viz)
```

# Regex

To search for an exact pattern, simply search for the pattern.

```{r, eval = FALSE}
# detect the name Charlie
x %>% str_detect("Charlie")
```

## Character Classes

Rather than searching for a specific character, the regex engine can search for a class of characters. There are three classes or groups of characters in which there are special regex expressions for. 

pattern | matches
--------|---------
\d      | digit character
\w      | word character (letter, digit, '_')
\s      | space character (space, tab, new line)

## Metacharacters

To allow for greater flexibility, there are a number of metacharacters that convey special meaning in regex pattern matching.

```{r metachar examples, echo = FALSE}
pattern <- c(".", "|", "(...)", "[...]", "{...}", "*", "+", "^", "$")
match <- c("any single character", "or (matches preceding or succeding string)", "used for backreference", "customized group of characters", "repetition quantifier", "matches previous character 0 or more times", "matches previous character 1 or more times", "matches at the beginning of a string", "matches at the end of a string")
data.frame(pattern, match) %>% set_colnames(c("pattern", "matches")) %>% 
  kable()
```

Here are further specifications on several of the characters listed above.

**[...]**

`[...]` can be used to select from a customized group of characters. To specify characters, list them out inside the brackets. To not match a customized group of characters, add a negation character like so: `[^...]`. 

To specifically search for only uppercase letters, use `[A-Z]`; for lowercase letters `[a-z]`. To search for both uppercase and lowercase letters, use `[A-Za-z]`.

**{...}**

To match a character (or a group of characters) multiple times, repetition quantifiers come in handy: `{...}`, `*`, or `+`. 

pattern | matches
--------|---------
{n}     | matches previous character exactly n times
{n,}    | matches previous character n or more times
{n, m}  | matches previous character n to m times
*       | matches previous character 0 or more times
+       | matches previous character 1 or more times

**(...)**

The metacharacters `(...)` are very useful in replacing strings. When wrapped around a string, it "remembers" the matched pattern. Then when specifying the replacement, it allows us to refer to the matched pattern by using `\n` where n refers to the nth wrapped expression. 

We can look at this example in R:
```{r example data}
x <- "Flintstone,Fred  *111-222-3333 fflinstone@gmail.com crane operator"
pattern <- "(\\w+)[, ]+(\\w+)\\s+([\\d-*]+)\\s+([\\w@.]+)\\s+(.+)"
## 1st () references the last name
## 2nd () references the first name
## 3rd () references the phone number
## 4th () references the email address
## 5th () references the person's occupation
x %>% str_replace(pattern = pattern, replacement = "\\2 \\1 is a \\5")
```

**^ and $**

The metacharacters `^` and `$` are used to match patterns at specific positions of a string. 
```{r, eval = FALSE}
grep "^\d"
grep "\w$"
```

**Using Metacharacters as Normal Characters**

To use these characters as regular characters, precede the character with escape characters or wrap then in square brackets. For example, to search for the character `+`, type `\+` or `[+]` into the search pattern string. 

# In R

In order to match these special regex expressions in R, type the backslashes twice, like so:

```{r, eval = FALSE}
# count the number of space characters in each string of x
x %>% str_count("\\s")
```

function | description
---------|--------------
`str_to_upper()`, `str_to_lower()`, `str_to_title()` | change string case
`str_dup()` | duplicate and concatenate character vector
`paste()` | join multiple strings into single string
`nchar()` | number of characters
`str_count()` | count number of matches in string
`str_sub()` | extract and replace substrings
`str_subset()` | keep strings matching a pattern
`str_detect()` | detect presence of a pattern
`str_locate()`, `str_locate_all()` | locate positions of patterns
`str_extract()`, `str_extract_all()` | extract pieces that match pattern
`str_match()`, `str_match_all()` | extract matched groups
`str_replace()`, `str_replace_all()` | replace matched patterns
`str_replace_na()` | replace NA
`str_split()` | split by pattern
`str_trim()` | trim white space
`str_pad()` | pad string
`str_wrap()` | wrap strings into formatted paragraphs
`word()` | extract words from sentence


