---
layout: post
title: "GLM: Poisson Regression"
date: "November 27, 2015"
categories: Statistics
tags: Regression
---

* TOC
{:toc}

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(jn.general)
lib(data, viz)
knitr::opts_chunk$set(fig.width = 5, fig.height = 5, fig.align = 'center')
library(faraway)
data(solder)
```

# Count Response
Suppose the response $$Y_i \sim Pois(\mu_i)$$ where $$\mu > 0$$ and the $$Y_i$$ are independent

$$P(Y_i = y_i) = \frac{e^{\mu_i} \mu_i^{y_i}}{y_i!}$$

Then

$$E(Y_i) = Var(Y_i) = \mu_i$$

The deviance is

$$ D = 2 \sum^n_{i = 1} \big[ y_i \log \left( \frac{y_i}{\hat{\mu}_i} \right)  - (y_i - \hat{\mu}_i) \big] $$

There few problems when the response variable has a Poisson distribution. 

* $$Var(Y) = \mu$$, which is not constant
* $$\mu > 0$$: but OLS may predict values less than zero

To alleviate these concerns, use a link function $$\eta_i = g(\mu) = \log(\mu)$$ such that $$g^{-1}(\eta) > 0$$ $$\forall \eta$$.

Several instances can be classified as Poisson:

* Count is some number out of some possible total, response is binomial. Poisson is good approximation if success probability is small
* If the probability of an event occuring in a given time interval is proportional to the length of that time interval and independent of other events. The number of events in any specified time interval is Poisson
* When the time between events is independent and identically exponentially distributed. The count of events in a given time period is Poisson

# Poisson Regression Model

## Modeling Counts
Fit the model

$$ \log(\mu) = \beta_0 + \beta_1 x_1 + ... + \beta_k x_k$$

In R, this model is fit like so
```{r, eval = FALSE}
glm(counts ~ ., data = data, family = poisson)
```

## Modeling Ratios

If the response involves ratios and the proportions are small, Poisson regression can be used (rather than logistic regression). Fit the model <br>
$$\log \left( \frac{count}{total} \right) = \beta_0 + \beta_1 x_1 + ... \beta_k x_k$$ <br>
$$ \log(count) = \log(total) + \beta_0 + \beta_1 x_1 + ... \beta_k x_k $$

In R, this model is fit like so
```{r, eval = FALSE}
glm(counts ~ offset(log(total)) + ., family = poisson, data = data)
```
The function `offset()` means to not fit a coefficient to a term (or set the coefficient to 1). In this model $$total = cells$$.

In SAS, fit these models like so
```{r, eval = FALSE}
proc genmod;
  model y = x1 x2 / type3 dist = poi link = log offset = total;
run;
```


# Beta Coefficients

## Interpretation of Coefficients
Interpretation of coefficients is similar to the logistic regression case. 

------------------------------|---------------------
$$\log(y \vert x_1 = x + 1)$$ | $$ = \beta_0 + \beta_1 (x + 1) + ... + \beta_k x_k$$
$$\log(y \vert x_1 = x) $$    | $$ = \beta_0 + \beta_1 x + ... + \beta_k x_k$$

Then 

-----------------------------------------------------------|---------------------
$$ \log(y \vert x_1 = x + 1) - \log(y \vert x_1 = x) $$ | $$ = \beta_1 $$
$$ \log \left( \frac{y \vert x_1 = x + 1}{y \vert x_1 = x} \right)$$ | $$ = \beta_1 $$
$$ \frac{y \vert x_1 = x + 1}{y \vert x_1 = x} $$          | $$ = e^{\beta_1} $$
$$ (y \vert x_1 = x + 1) $$                                | $$= e^{\beta_1} * (y \vert x_1 = x)$$

Holding all other predictors constant, a unit increase in $$x_i$$ increases the count by a factor of $$exp(\hat{\beta}_i)$$

# Overdispersion
Overdispersion occurs in poisson regression when $$Var(Y_i)$$ is greater than the assumed $$Var(Y_i) = \mu_i $$. With Poisson models, overdispersion may be visually assessed by plotting $$(y - \hat{\mu})^2$$ vs $$\mu$$ with the identity line. Overdispersion would result in a plot where the majority of points are above the identity line, indicating $$Var(Y) > E(Y)$$. 

The method to adjust for overdispersion involves multiplying the variance by a factor $$\sigma^2$$ to obtain <br>
$$Var(Y_i)^* = \sigma^2 \mu_i $$

The steps for assessing overdispersion are listed in the [GLM Testing and Diagnostics: Overdispersion][glm_diagnostics_post]{:target = "blank"}.

## Negative Binomial Regression
One alternative to the Poisson regression with overdispersion is negative binomial regression. The response $$Z$$ is assumed to follow a negative binomial distribution with $$Z$$ being the number of trials until the $$k^{th}$$ success 

$$P(Z = z) = \left(\begin{array}
{rrr}
z - 1\\
k - 1
\end{array}\right)
p^k (1 - p)^{z - k}
$$

where $$ z > k$$, $$p$$ is the probability of success.

This equation can be rewritten; let $$Y = Z - k$$ and $$p = (1 + \alpha)^{-1}$$

$$P(Y = y) = \left(\begin{array}
{rrr}
y + k - 1\\
k - 1
\end{array}\right)
\frac{\alpha^y}{(1 + \alpha)^{y + k}}$$

where $$E(Y) = k\alpha = \mu$$ and $$Var(Y) = k\alpha + k\alpha^2 = \mu + D\mu^2$$. The parameter $$D = 1/k$$ is called the dispersion parameter.

The negative binomial regression model is

$$\eta = \log \left( \frac{\mu}{\mu + k} \right) = x \beta$$

Greater heterogeneity in Poisson resulting in overdispersion results in a larger value of $$D$$. As $$D \rightarrow 0$$, $$Var(Y) \rightarrow \mu$$ and the negative binomial and Poisson regression gives the same inference. 

In R, fit a negative binomial function with
```{r, message = FALSE}
# load package
library(MASS)

# fit model
mod <- glm.nb(skips ~ ., solder)
summary(mod)
```
Here the value $$\hat{k} = 4.4$$ with a standard error of 0.5. 

In SAS, fit models using
```{r, eval = FALSE}
proc genmod;
  class x1 (param=ref ref=first);
  model y = x1 x2 / type3 dist=negbin;
run;
```


[glm_diagnostics_post]: http://jennguyen1.github.io/nhuyhoa//2015/11/GLM-Testing-and-Diagnostics.html#overdispersion
