---
layout: post
title: "Mapping Values"
date: "October 10, 2015"
categories: ['data wrangling']
---

* TOC
{:toc}

# In R
```{r, echo = FALSE}
library(jn.general)
lib(data)
subjects <- c("reading", "writing", "language arts", "english", "algebra", "geometry", "calculus", "statistics", "science")
```

## ifelse
The function `ifelse()` provides an easy way to do this. The format is `ifelse(CONDITION, TRUE_VALUE, FALSE_VALUE)`.

Here's an example:
```{r}
# change the vowels to "vowel" and non-vowels to "consonant"
ifelse(letters %in% c("a", "e", "i", "o", "u"), "vowel", "consonant")

# change the vowels to "vowel" and keep non-vowels as they are
ifelse(letters %in% c("a", "e", "i", "o", "u"), "vowel", letters)
```

## mapvalues
The function `plyr::mapvalues()` is a more versatile version of `ifelse()`. The format is `mapvalues(VECTOR, FROM_VALUES, TO_VALUES)`, where the $$i^{th}$$ value in FROM_VALUES is changed to the $$i^{th}$$ value in TO_VALUES.

Here's an example:
```{r}
# start with this vector
subjects

# simplify subjects to reading and math
mapvalues(subjects, 
          c("reading", "writing", "language arts", "english",
            "algebra", "geometry", "calculus", "statistics"),
          c(rep("read", 4), rep("math", 4)))
```

Here the reading-related subjects were simplified to `"read"` and the mathematics-related subjects to were simplified to `"math"`. Notice how values not listed, such as `"science"`, was not affected. 

## setnames
The function `data.table::setnames()` is equivalent to `mapvalues()` but is used solely for renaming data frame columns.

Here's an example:

<div class = "dftab">
```{r, echo = FALSE}
iris %>% nhuyhoa_df_print()
```
</div><p></p>

```{r}
# make the column names lower case
setnames(iris, colnames(iris), tolower(colnames(iris)))
```

<div class = "dftab">
```{r, echo = FALSE}
iris %>% nhuyhoa_df_print()
```
</div><p></p>

## Vectorize
Sometimes you may just want to write your function that uses conditionals to relate certain values. Your function inputs may just take one data structure (perhaps one value) and may throw an error if you pass in something along the lines of a vector. 

For these instances, there is a function, called `Vectorize()` that can transform these types of functions so that it will work with vectors. 

{% highlight r %}
# create a function f
f <- function(x){...}

# vectorize the function so it can take a vector as input
vf <- Vectorize(f)
{% endhighlight %}



