---
layout: post
title: "GLM: Understanding Output"
date: "November 25, 2015"
categories: statistics
---

* TOC
{:toc}

# Summary Table
```{r, echo = FALSE, message = FALSE}
library(jn.general)
lib(data, viz)
knitr::opts_chunk$set(fig.width = 5, fig.height = 5, fig.align = 'center')
library(faraway)
data(orings)
```

The output from glms in R are basically the same regardless of the link. Here we will look at an example output. 

```{r}
mod <- glm(cbind(damage, 6 - damage) ~ temp, data = orings, family = binomial)
summary(mod)
```

* Coefficient table: $$\hat{\beta}$$ estimates and standard errors, $$\hat{\beta}$$ are normally distributed
* Model Fit Statistics: deviances and AIC, deviances have $$X^2$$ distribution
* Fischer Scoring iterations: number of iterations in the IRWLS algorithm

# Beta Confidence Intervals
The $$100(1 - \alpha)$$% confidence interval for $$\hat{\beta}_i$$ is 

$$\hat{\beta}_i \pm z_{\alpha /2} se(\hat{\beta}_i)$$

# Predicted Values and Prediction Confidence Intervals
For a given set $$ x_0 $$, we can predict the reponse.

Procedure:

* Compute point estimate: $$ \hat{\eta} = x_0 \hat{\beta} $$
* From R we can extract the variance matrix of the coefficients: `m$cov.unsealed`
* Compute variance: $$ var(\hat{\eta}) = x'_0 (X'WX)^{-1} x_0$$
* Compute confidence interval: $$ exp \left( \hat{\eta} \pm z_{\alpha/2} \sqrt{var(\hat{\eta})} \right)$$

We can also use built in R functions:

* Predict $$\hat{\eta}$$: `predict(object, newdata, type = "link")`
* Predict $$\hat{\mu}$$: `predict(object, newdata, type = "response")`