---
title: "To Be or Not To Be: Subsetting"
layout: post
categories: r data_wrangling
---

* TOC
{:toc}

```{r global_opts, echo = FALSE}
library(jn.general)
lib(viz)
```

# Subsetting Functions in R

In a previous post, I dicussed the subsetting functions available via the dplyr package, including the `filter()`, `slice()`, and `select()` functions. There are a number of additional ways to subset data frames in R. 

```{r subsetting functions, echo = FALSE}
functions <- c("head()", "tail()", "magrittr::extract()", "dplyr::sample_frac()", "dplyr::sample_n()", "dplyr::distinct()", "jn.general::duplicated_data()")
purpose <- c("grabs first n rows", "grabs last n rows", "wrapper for '['; takes numeric or character arguments and return corresponding rows/columns", "randomly selects a given fraction of rows", "randomly selects a given number of rows", "subsets to distinct values of a data frame, given a specific variable", "returns all duplicated rows by a given variable")
data.frame(functions, purpose) %>% kable(format = "html")
```
<p></p>

However what if rather than tossing the data that does not correspond to the subset, we wanted to keep it? Say perhaps, we want to subset, perform a sequence of operations, and then somehow relate the two to each other. 

One option would be run the subsetting function two times. Once with the conditions that's required, and then again with the converse argument. 
```{r, eval = FALSE}
mtcars %>% subset( gear != 4 )
mtcars %>% subset( gear == 4 )

mtcars %>% dplyr::select(one_of("mpg", "cyl", "hp"))
mtcars %>% dplyr::select(-one_of("mpg", "cyl", "hp"))
```

This tends to work pretty well when the filtering condition is pretty simple. However, when we get into multiple arguments, negation can get pretty ugly fast. 

In other cases, obtaining the "anti-subset" is not all that straight forward. For example, if a subset of the data was obtained randomly, as with `sample_n()` or `sample_frac()`, it's difficult to find the portion of the data that wasn't returned by the function. Another example of not being able to easily negate is with `duplicated_data()`, which returns all duplicated rows by a given variable. 

Having faced this issue many times, I wrote the `to_be()` function as a way to obtain both the subsetted data and the "anti-subset". 

# To Be or Not To Be

The function`to_be()` is quite simple when broken down. It takes in a data frame, a subsetting function, and any additional arguments needed for the subset. After performing the needed operations, `to_be()` returns a list of two objects. The first object returned, "to_be", is the resulting data frame when the subset function is applied to the input data. The second object returned, "not_to_be" is the anti-subset, obtained by considering a total of three cases: subset by rows, subset by columns, and subset by rows and columns. 

## Subset by Rows

```{r subset by rows ex}
by_row <- to_be(mtcars, dplyr::filter, cyl != 4, mpg > 20)
```

{% highlight r %}
# data corresponding to cars where the cyl is not 4 and mpg is greater than 20
by_row[["to_be"]]
{% endhighlight %}
<div class = "dftab">
```{r, echo = FALSE}
by_row[["to_be"]] %>% kable(format = "html")
```
</div><p></p>

{% highlight r %}
# data corresponding to cars where cyl is 4 or mpg is less than 20
by_row[["not_to_be"]] %>% head
{% endhighlight %}
<div class = "dftab">
```{r, echo = FALSE}
by_row[["not_to_be"]] %>% nhuyhoa_df_print()
```
</div>

## Subset by Columns

```{r subset by col ex}
by_col <- to_be(mtcars, dplyr::select, -one_of(c("gear", "vs")))
```


{% highlight r %}
# data corresponding to all car information except gear & vs 
by_col[["to_be"]] %>% head
{% endhighlight %}
<div class = "dftab">
```{r, echo = FALSE}
by_col[["to_be"]] %>% nhuyhoa_df_print()
```
</div><p></p>

{% highlight r %}
# data corresponding to gear and vs information of cars
by_col[["not_to_be"]] %>% head
{% endhighlight %}
<div class = "dftab">
```{r, echo = FALSE}
by_col[["not_to_be"]] %>% nhuyhoa_df_print()
```
</div>


## Subset by Rows and Columns

```{r subset by both ex}
by_both <- to_be(mtcars, subset, subset = mpg > 20, select = c(cyl, hp, wt))
```

When the function both subsets rows and selects columns, the anti-subset data frame will replace the extracted values with NAs. 

{% highlight r %}
# data corresponding to cyl, hp, wt of cars with an mpg greater than 20
by_both[["to_be"]]
{% endhighlight %}
<div class = "dftab">
```{r, echo = FALSE}
by_both[["to_be"]] %>% kable(format = "html")
```
</div><p></p>

{% highlight r %}
# data corresponding to cyl, hp, wt of cars with an mpg greater than 20 are replaced with NAs
by_both[["not_to_be"]] 
{% endhighlight %}
<div class = "dftab">
```{r, echo = FALSE}
by_both[["not_to_be"]] %>% kable(format = "html")
```
</div>

# Challenge Round
Let’s do something a little more challenging to see to_be in action. With this data set, we want to remove the duplicated id entries by hand and merge them back into the non-duplicates. 

<div class = "dftab">
```{r challenge data, echo = FALSE}
set.seed(1)
data <- rdata(10,
  gen_seq(1, 10, name = "id"),
  gen_age(),
  gen_gender(),
  gen_race(),
  gen_state(),
  .id = FALSE
) %>% arrange(id)

data %>% kable(format = "html")
```
</div><p></p>

```{r find duplicates}
result <- to_be(data, duplicated_data, id)
```

The duplicated id entries:

{% highlight r %}
result[["to_be"]]
{% endhighlight %}
<div class = "dftab">
```{r to be, echo = FALSE}
x <- result[["to_be"]]
rownames(x) <- NULL
x %>% kable(format = "html")
```
</div><p></p>

The non-duplicated id entries:

{% highlight r %}
result[["not_to_be"]]
{% endhighlight %}
<div class = "dftab">
```{r not to be, echo = FALSE}
y <- result[["not_to_be"]] 
rownames(y) <- NULL
y %>% kable(format = "html")
```
</div><p></p>

From the duplicated id's, I'll keep the id that corresponds to a female or, if none of the duplicated entries are female, an id corresponding to someone greater than 50 years old. 

{% highlight r %}
result[["to_be"]] %>% group_by(id) %>% arrange(gender, desc(age)) %>% slice(1)
{% endhighlight %}
<div class = "dftab">
```{r, echo = FALSE}
z <- x %>% group_by(id) %>% arrange(gender, desc(age)) %>% slice(1) 
z %>% kable(format = "html")
```
</div><p></p>

Now combine the results back together.

{% highlight r %}
result[["to_be"]] %>% 
  group_by(id) %>% 
  arrange(gender, desc(age)) %>% 
  slice(1) %>% 
  rbind(result[["not_to_be"]])
{% endhighlight %}
<div class = "dftab">
```{r, echo = FALSE}
a <- result[["to_be"]] %>% group_by(id) %>% arrange(gender, desc(age)) %>% slice(1) %>% rbind(result[["not_to_be"]]) %>% arrange(id)
rownames(a) <- NULL
a %>% kable(format = "html")
```
</div>
