---
layout: post
title: "Hypothesis Testing in Regression"
date: "October 23, 2015"
categories: statistics
---

* TOC
{:toc}

```{r, echo = FALSE}
library(jn.general)
```

# F Test
Suppose we want to compare 2 models, where one is nested in the other. Let H correspond to the simplified model with p - q parameters. Let A correspond to the full model with p parameters. Then we can compare the two models with an F test.

$$ F = \frac{(SSE_H - SSE_A)/ (df_A - df_H)}{SSE_A/df_A} $$

where $$ F $$ ~ $$ F_{q, n - p} $$ and SSE is the sum square errors and DF is the residual degrees of freedom corresponding to the specified model.

# Example
```{r}
set.seed(1)
y <- rnorm(50)
x <- runif(50)
z <- rexp(50, 1.5)

# by hand
modh <- lm(y ~ x)
moda <- lm(y ~ x + z)
RSSh <- anova(modh)["Sum Sq"][2,1]
RSSa <- anova(moda)["Sum Sq"][3,1]

F <- (RSSh - RSSa) / (1) / (RSSa/47)
pf(F, 1, 47, lower.tail = FALSE)

# using built in R
anova(modh, moda)
```

Thus we conclude that for this data set, the full and simplified model are not significantly different frome each other. Thus the "best" model is the simplified model.