---
layout: post
title: "Exception Handling"
date: "August 21, 2017"
categories: Software
tags: Pipelines
---

* TOC
{:toc}

```{r, echo = FALSE}
library(jn.general) 
lib(data)
```

# Exception Handling 

**R**

The package `purrr` has 3 functions to deal with error handling. These functions are applied to other functions and can handle error output.

* `safely()` returns a list with 2 elements: result and error, one of which is always NULL
* `quietly()` returns a list with 4 elements: result, output, warnings, and messages
* `possibly()` replaces errors in 1 object with a default value

```{r}
map(list(1, 2, "hi"), safely(log))
map(list(1, 2, "hi"), quietly(log))
map(list(1, 2, "hi"), possibly(log, NA_real_))
```


Similar to other languages `tryCatch()` can be used to deal with how error, warning, and messages are dealt with. 
```{r}
error_handling <- function(code){
  tryCatch(
    code, 
    error = function(e){"error"}, 
    warning = function(e){"warning"},  
    message = function(e){"message"}
  )
}

error_handling(stop("!"))
```


**Python**

Exceptions can be raised by doing 
{% highlight python %}
raise Exception("")
{% endhighlight %}

There are a variety of built in exceptions that you can use or you may write your own.

{% highlight python %}
class Error(Exception):
  def __init__(self, message, var2):
    self.message = message
    self.var2 = var2
{% endhighlight %}

Try/catch is also available in python.

{% highlight python %}
try:
  # OPERATIONS
except (Exception1, Exception2) as e:
  # SPECIFIED ERROR
except:
  # ALL OTHER ERRORS
else:
  # RUNS IF NO EXCEPTIONS IN TRY
finally:
  # OPERATIONS
{% endhighlight %}

Possible excepions

* IOError
* IndexError

# Assertions

Within functions, assertions can be used to detect errors.

**R**

```{r, eval = FALSE}
f <- function(){
  assertthat::assert_that(condition, msg = "Error message")
}
```

**Python**

{% highlight python %}
def f():
  assert condition, error_msg
{% endhighlight %}

# Unit Testing

**R**

Unit tests in R can be done with the package `testthat`. 

```{r, eval = FALSE}
context("Script Purpose")

testthat::test_that("Testing set 1", {
  testthat::expect_Blah(...)
})

testthat::test_that("Testing set 2", {
  testthat::expect_Blah(...)
})
```


To initiate, run the function `devtools::use_testthat()`. It will create all the necessary files. Test scripts can be placed in the folder *tests/testthat*. 

**Python**

{% highlight python %}
import unittest
import my_functions # import your own code

class TestName(unittest.TestCase):

  # setup any objects needed for testing
  def setUp(self):
    pass

  # any series of tests
  def test_name(self):
    self.assertBlah(my_functions.f(), ...)
    
  # cleanup any objects needed for testing
  def tearDown(self)
  
  
# call tests  
if __name__ == '__main__':
  unittest.main()
{% endhighlight %}

{% highlight bash %}
# run tests
python test_unittest.py -v
{% endhighlight %}


Here are a list of common assertions

* `assertEqual(a,b)`, `assertNotEqual()`
* `assertTrue(x)`, `assertFalse()`
* `assertIs(a,b)`, `assertIsNot()`
* `assertNone(x)`, `assertIsNotNone()`
* `assertIn(a,b)`, `assertNotIn()`
* `assertIsInstance(a,b)`, `assertNotIsInstance()`
* `assertAlmostEqual(a,b,places,delta)`, `assertNotAlmostEqual()`
* `assertGreater(a,b)`, `assertGreaterEqual()`
* `assertLess(a,b)`, `assertLessEqual()`
* `assertRegex(t,regexp)`, `assertNotRegex()`
* `assertCountEqual(a,b)`
* `assertListEqual(a,b)`, `assertTupleEqual()`, `assertSetEqual()`, `assertDictEqual()`
* `assertRaises()`, `assertRaisesRegex()`

These methods have a parameter `msg` in which you can provide additional messages upon fail.

It is a good idea to unit test modules and packages that you use repetitively. For one-off scripts (cleaning/transforming data, etc), unit testing is unecessary but you can sprinkle asserts throughout the code. 

# Debugging

**R**

When placed inside of a function, `browser()` will trigger a debugging session in interactive mode. 


